Trends of SDR in Thessaloniki

### 2023-09-12 Article.Rmd
Commit 2023-09-12 20:14

<!-- adjusted Haurwitz -->
\begin{equation}
\text{SDR}_\text{CSref} = a \times \text{SDR}_\text{Haurwitz} = `r signif(alpha,3)` \times 1098 \times \cos( \text{SZA} ) \times \exp \left( \frac{ - 0.057}{\cos(\text{SZA})} \right) \label{eq:ahau}
\end{equation}

where: $\text{SDR}_\text{CSref}$ is the reference clear sky SDR, in $\text{w}\,\text{m}^{-2}$.

@@

We calculated the SDR trends, from the mean daily values of the departure from the seasonal values, for the three sky conditions (Table\ \ref{tab:trendtable}).
We present only the plot for the trend under all-sky conditions (Figure\ \ref{fig:trendALL}), as the plots for clear-sky condition and cloud-sky condition, are very similar.
In the studied time period, there is no significant break or change in the variability of the overall trends.
Other studies for the European region are reporting a change of the SDR slope, around the 1980, few years before the start of our records [@Wild2021; @Yuan2021; @Ohmura2009]. It is interesting to note, that for the observation period, the trend of the TSI is
$`r format(TSIslopeAll[1,1], scientific = T, digits = 2)`\%/year$, and thus we can not attribute any major effect on SDR trend to Sun variability.

@@

```{r trendALL, echo=F, out.width='.70\\linewidth', fig.cap=paste0("Anomaly (\\%) of the daily SDR relative to climatological values for ", year(min(ALL_1_daily_DESEAS$Date)), " - ", year(max(ALL_1_daily_DESEAS$Date)), ". The black line shows the long term trend for all-sky conditions.")}
knitr::include_graphics("./images/LongtermTrends-2.png")
```

Although the trends are quite constant, using plots of the cumulative sums we can reveal some structure within the trends and some differences between the sky conditions [@Regier2019]. 
On all three plots on Figure\ \ref{fig:cumsum}, we observe a downward part at the start of the period until about 2000, an almost steady part until about 2016 for All-sky and Clear-sky conditions (Figures\ \ref{fig:cumsum-1}, \ref{fig:cumsum-2}) and until 2009 for Cloud-sky conditions (Figure\ \ref{fig:cumsum-3}), and finally, a steep upward part until the present.
For a uniform trend, we expected the cumulative sum plot of the anomaly to have a symmetric 'V' shape.
This would indicate, that the anomaly is evenly distributed around the climatological mean.
In our case, for all sky conditions, we observe periods where the cumulative sum remains relative steady, roughly in the middle of the study dataset.
Another distinct feature of the cumulative plots, is the difference, between the Cloud-sky and the rest of the sky conditions, where they display a supplementary behavior. 
The causes for these discrepancies haven't been investigated yet.

<!--
This can make us to suspected that clear-Sky conditions SDR and cloud-sky SDR is affected by different mechanisms.
The similarity of the All-sky and Clear-sky conditions, can make us to suspected that on both sky conditions the attenuation of SDR has .
-->

```{r cumsum, echo = FALSE, fig.cap='Running cumulative sum plots of the daily SDR for three different sky conditions.', fig.subcap=c('All-sky conditions.', 'Clear-sky conditions.', 'Cloud-sky conditions.'), out.width='.32\\linewidth', fig.asp=1, fig.ncol = 3}
knitr::include_graphics(c(
    "./images/CumulativeDailyCumSum-1.png",    # ALL sky
    "./images/CumulativeDailyCumSum-5.png",    # CLEAR sky
    "./images/CumulativeDailyCumSum-9.png"))   # Cloud sky
```

@@

<!-- ```{r trendCLEAR, echo=F, fig.cap="Daily SDR seasonal anomaly and long term trend, for clear sky conditions."} -->
<!-- knitr::include_graphics("./images/LongtermTrends-4.png") -->
<!-- ``` -->

<!-- ```{r trendCLOUD, echo=F, fig.cap="Anomaly (\\%) of the daily SDR relative to climatological values for 1993 - 2023. The black line shows the long term trend for cloud-sky conditions."} -->
<!-- knitr::include_graphics("./images/LongtermTrends-6.png") -->
<!-- ``` -->

@@

The relative position of the Sun in the sky, is a major drive of the attenuation of SDR reaching the ground, as it is affected by the radiation path in the atmosphere and possible human activities in urban environment emitting aerosols [@Wang2021].
In order to estimate the contribution of the SZA, we group the anomaly data in $1^\circ$ SZA bins, and calculated the overall trend for each bin before noon and after noon (Figure\ \ref{fig:szatrends}).
Although there is a seasonal variation of the minimum SZA, that is not examined here.
The brightening effect of SDR is stronger for All-sky and Clear-sky condition for low Sun elevation angles (Figures\ \ref{fig:szatrends}a and \ref{fig:szatrends}b).
In the case of Cloud-sky condition (Figure\ \ref{fig:szatrends}c), there is no strong correlation between SDR brightening and SZA, except for some low Sun angles, that occur only for a small part of the year, and thus we can not draw more conclusions from this analysis for the overall trends.

@@

Similar to the long term trends discussed above, we have produced the deseasonalized trend for the different sky conditions and for each season of the year, using the corresponding mean monthly values
(Figure\ \ref{fig:seasonalALL} and Table\ \ref{tab:trendseasontable}).
For All-sky condition the brightening of SDR is very significant for the Winter, close to the long term trend for the Autumn, and almost four times lower than the long term trend for Sprint and Summer, indicating a relation to the cloud pattern of each season.
Moreover, when examining the Clear- and Cloud-sky conditions the trends between these conditions are almost complementary to each other for Winter and Autumn respectively. On the latter cases, during Spring and Summer the statistical significance along with the actual trend are very low for a meaningful comparison.

@@

<!-- ```{r seasonalCLEAR, echo=F, fig.cap="Trends by season for Clear-sky condition. Displays monthly means of daily means."} -->
<!-- knitr::include_graphics("./images/SeasonalTrends-2.png") -->
<!-- ``` -->

@@

<!-- \footnotesize -->
```{r trendseasontable, echo=F}
pp <- read.csv("./figures/tbl_longterm_trends_season.dat",
               strip.white = TRUE,
               sep         = ";",
               header      = TRUE)
pp <- pp[grep("SDR" ,pp$var), ]
pp$slope.sd           <- NULL
pp$slope.ConfInt_0.95 <- NULL
pp$slope.ConfInt_0.99 <- NULL
pp$var                <- NULL
rownames(pp)          <- NULL
names(pp) <- c("Trend [%/year]", "Trend p-value", "Sky condition", "Season", "Statistical signif. [%]")
## order by season
target    <- c("Winter", "Spring", "Summer", "Autumn")
new_order <- sapply(target, function(x,pp){which(pp$Season == x)}, pp = pp)
pp        <- pp[new_order,]
## put all sky first
A  <- which(pp$`Sky condition` == "All sky cond.")
B  <- which(pp$`Sky condition` != "All sky cond.")
pp <- pp[c(A,B),]
## numeric precision
pp$`Trend [%/year]`          <- signif(pp$`Trend [%/year]`, 3)
pp$`Trend p-value`           <- signif(pp$`Trend p-value`,  3)
pp$`Statistical signif. [%]` <- signif(pp$`Statistical signif. [%]`,  3)

@@

We demonstrate that in the period 
`r paste(year(min(ALL_1_daily_DESEAS$Date)), " - ", year(max(ALL_1_daily_DESEAS$Date)))`, there is a brightening trend of SDR for All-sky conditions ($`r alltrend`\%/year$).
On a previous study [@Bais2013], for the period 1993 - 2011 the trend was found to be $0.33\%/year$.
The increase of this trend, indicates that the phenomenon and probably the causes, are still occurring.
Accordingly, we found a similar trend for Clear-sky conditions ($`r cleartrend`\%/year$) that further supports that the brightening can be independent of the cloud occurrences. 
We have calculated the effect of TSI variability to be a negligible factor in SDR variability.
Unfortunately, there is no available data for this period for the Aerosols, in order to estimate their effect on SDR, which is the most probable main cause.
The attenuation of SDR by Aerosols over Europe have been proposed as major factor by @Wild2021.
The dimming effect of SDR under Cloud-sky condition ($`r cloudtrend`\%/year$), shows that clouds can also attenuate the amount of SDR reaching the ground, and changes on their occurrences can effect the net energy balance.
Because we have no adequate data to investigate long term changes of cloud formation in the region, we can not infer if the SDR trend we observe under Cloud-sky is biased by changes in clouds and local climate.
The observed brightening of SDR over Thessaloniki, is correlated with the SZA (higher SZA, have stronger brightening trends). And also, with the seasons of the year, where Winter and Autumn show a significant strong trend, in contrast to Spring and Summer.

@@

>>>>

<!-- adjusted Haurwitz -->
\begin{equation}
\text{SDR}_\text{CSref} = a \times \text{SDR}_\text{Haurwitz} = `r signif(alpha,3)` \times 1098 \times \cos(\theta) \times \exp \left( \frac{ - 0.057}{\cos(\theta)} \right) \label{eq:ahau}
\end{equation}

where: $\text{SDR}_\text{CSref}$ is the reference clear sky SDR, in $\text{w}\,\text{m}^{-2}$ and $\theta$ is the solar zenith angle (SZA).

@@

We calculated the linear SDR trends, from the departures of the mean daily values from the seasonal values, and for the three sky conditions (Table\ \ref{tab:trendtable}).
In Figure\ \ref{fig:trendALL} we present only the time series under all-sky conditions; the plots for clear-sky and cloud-sky conditions, are very similar.
In the studied time period, there is no significant break or change in the variability of the overall trends.
Other studies for the European region are reporting a change of the SDR slope, around the 1980, few years before the start of our records [@Wild2021; @Yuan2021; @Ohmura2009]. It is interesting to note, that for the observation period, the trend of the TSI is
$`r format(TSIslopeAll[1,1], scientific = T, digits = 2)`\%/year$, and thus we can not attribute any major effect on SDR trend to Sun variability.

@@

```{r trendALL, echo=F, out.width='.70\\linewidth', fig.cap=paste0("Anomaly (\\%) of the daily All-sky SDR, relative to climatological values for ", year(min(ALL_1_daily_DESEAS$Date)), " - ", year(max(ALL_1_daily_DESEAS$Date)), ". The black line shows the long term trend for all-sky conditions.")}
knitr::include_graphics("./images/LongtermTrends-2.png")
```

Although the trends are quite constant, using plots of the cumulative sums we can reveal some structure within the trends and some differences between the sky conditions [@Regier2019]. 
On all three plots on Figure\ \ref{fig:cumsum}, we observe a downward part at the start of the period until about 2000, an almost steady part until about 2016 for All-sky and clear-sky conditions (Figures\ \ref{fig:cumsum-1}, \ref{fig:cumsum-2}) and until 2009 for cloud-sky conditions (Figure\ \ref{fig:cumsum-3}), and finally, a steep upward part until the present.
For a uniform trend, we expected the cumulative sum plot of the anomaly to have a symmetric 'V' shape.
This would indicate, that the anomaly is evenly distributed around the climatological mean.
In our case, for all sky conditions, we observe periods where the cumulative sum remains relative steady, roughly in the middle of the study dataset.
Another distinct feature of the cumulative plots, is the difference, between the cloud-sky and the rest of the sky conditions, where they display a supplementary behavior. 
The causes for these discrepancies haven't been investigated yet.

<!--
This can make us to suspected that clear-Sky conditions SDR and cloud-sky SDR is affected by different mechanisms.
The similarity of the All-sky and clear-sky conditions, can make us to suspected that on both sky conditions the attenuation of SDR has .
-->

```{r cumsum, echo = FALSE, fig.cap='....... Running cumulative ...... sum plots of the daily SDR for three different sky conditions.', fig.subcap=c('All-sky conditions.', 'clear-sky conditions.', 'Cloud-sky conditions.'), out.width='.32\\linewidth', fig.asp=1, fig.ncol = 3}
knitr::include_graphics(c(
    "./images/CumulativeDailyCumSum-1.png",    # ALL sky
    "./images/CumulativeDailyCumSum-5.png",    # CLEAR sky
    "./images/CumulativeDailyCumSum-9.png"))   # Cloud sky
```

@@

@@

The relative position of the Sun in the sky, is a major drive of the attenuation of SDR reaching the ground, as it is affected by the radiation path in the atmosphere and possible human activities in urban environment emitting aerosols [@Wang2021].
In order to estimate the contribution of the SZA, we group the anomaly data in $1^\circ$ SZA bins, and calculated the overall trend for each bin before noon and after noon (Figure\ \ref{fig:szatrends}).
Although there is a seasonal variation of the minimum SZA, that is not examined here.
The brightening effect of SDR is stronger for All-sky and clear-sky condition for low Sun elevation angles (Figures\ \ref{fig:szatrends}a and \ref{fig:szatrends}b).
In the case of cloud-sky condition (Figure\ \ref{fig:szatrends}c), there is no strong correlation between SDR brightening and SZA, except for some low Sun angles, that occur only for a small part of the year, and thus we can not draw more conclusions from this analysis for the overall trends.

@@

Similar to the long term trends discussed above, we have produced the deseasonalized trend for the different sky conditions and for each season of the year, using the corresponding mean monthly values
(Figure\ \ref{fig:seasonalALL} and Table\ \ref{tab:trendseasontable}).
For All-sky condition the brightening of SDR is very significant for the Winter, close to the long term trend for the Autumn, and almost four times lower than the long term trend for Sprint and Summer, indicating a relation to the cloud pattern of each season.
Moreover, when examining the clear- and cloud-sky conditions the trends between these conditions are almost complementary to each other for Winter and Autumn respectively. On the latter cases, during Spring and Summer the statistical significance along with the actual trend are very low for a meaningful comparison.

@@

<!-- ```{r seasonalCLEAR, echo=F, fig.cap="Trends by season for clear-sky condition. Displays monthly means of daily means."} -->
<!-- knitr::include_graphics("./images/SeasonalTrends-2.png") -->
<!-- ``` -->

@@

<!-- \footnotesize -->
```{r trendseasontable, echo=F}
pp <- read.csv("./figures/tbl_longterm_trends_season.dat",
               strip.white = TRUE,
               sep         = ";",
               header      = TRUE)
pp <- pp[grep("SDR" ,pp$var), ]
pp$slope.sd           <- NULL
pp$slope.ConfInt_0.95 <- NULL
pp$slope.ConfInt_0.99 <- NULL
pp$var                <- NULL
rownames(pp)          <- NULL
names(pp) <- c("Trend [%/year]", "Trend p-value", "Sky condition", "Season", "Statistical signif. [%]")
## order by season
target    <- c("Winter", "Spring", "Summer", "Autumn")
new_order <- sapply(target, function(x,pp){which(pp$Season == x)}, pp = pp)
pp        <- pp[new_order,]
# ## put all sky first
# A  <- which(pp$`Sky condition` == "All sky cond.")
# B  <- which(pp$`Sky condition` != "All sky cond.")
# pp <- pp[c(A,B),]
## numeric precision
pp$`Trend [%/year]`          <- signif(pp$`Trend [%/year]`, 3)
pp$`Trend p-value`           <- signif(pp$`Trend p-value`,  3)
pp$`Statistical signif. [%]` <- signif(pp$`Statistical signif. [%]`,  3)

@@

We demonstrate that in the period 
`r paste(year(min(ALL_1_daily_DESEAS$Date)), " - ", year(max(ALL_1_daily_DESEAS$Date)))`, there is a brightening trend of SDR for All-sky conditions ($`r alltrend`\%/year$).
On a previous study [@Bais2013], for the period 1993 - 2011 the trend was found to be $0.33\%/year$.
The increase of this trend, indicates that the phenomenon and probably the causes, are still occurring.
Accordingly, we found a similar trend for clear-sky conditions ($`r cleartrend`\%/year$) that further supports that the brightening can be independent of the cloud occurrences. 
We have calculated the effect of TSI variability to be a negligible factor in SDR variability.
Unfortunately, there is no available data for this period for the Aerosols, in order to estimate their effect on SDR, which is the most probable main cause.
The attenuation of SDR by Aerosols over Europe have been proposed as major factor by @Wild2021.
The dimming effect of SDR under cloud-sky condition ($`r cloudtrend`\%/year$), shows that clouds can also attenuate the amount of SDR reaching the ground, and changes on their occurrences can effect the net energy balance.
Because we have no adequate data to investigate long term changes of cloud formation in the region, we can not infer if the SDR trend we observe under cloud-sky is biased by changes in clouds and local climate.
The observed brightening of SDR over Thessaloniki, is correlated with the SZA (higher SZA, have stronger brightening trends). And also, with the seasons of the year, where Winter and Autumn show a significant strong trend, in contrast to Spring and Summer.

@@

<div id="refs"></div>

\newpage
# Appendix

```{r trendCLEAR, echo=F, fig.cap=paste0("Anomaly (\\%) of the daily clear-sky SDR, relative to climatological values for", year(min(ALL_1_daily_DESEAS$Date)), " - ", year(max(ALL_1_daily_DESEAS$Date)), ". The black line shows the long term trend for clear-sky conditions.")}
knitr::include_graphics("./images/LongtermTrends-5.png")
```

```{r trendCLOUD, echo=F, fig.cap=paste0("Anomaly (\\%) of the daily cloud-sky SDR, relative to climatological values for ", year(min(ALL_1_daily_DESEAS$Date)), " - ", year(max(ALL_1_daily_DESEAS$Date)), ". The black line shows the long term trend for cloud-sky conditions.")}
knitr::include_graphics("./images/LongtermTrends-8.png")
```
